For Induction-Induction, Induction is Enough

### 2019-11-11 paper/paper.tex
part of the intro

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the James
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct this theory using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda.

\end{abstract}

@@

\todo{Note by Ambrus: to be beautified, I only wanted to add the line of
  thought.}

\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work finish, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)n}

\todo{
  András: construction of infinitary IITs doesn't (yet) work. Should we drop
  infinitary functions from signatures?
  }

Most mutual inductive types (e.g.\ isEven-isOdd) can be reduced to indexed
inductive types where the index gives the information which sort we
meant. (Indexed inductive types can be reduced to inductive types using a method
similar to what we use here, preterms and well-formedness predicate.)
Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual definition
of a type and a family of types over the first one. The above reduction does not
work anymore.

Examples of IITs: Con-Ty is the minimal interesting one, then we have
James Chapman style \cite{chapman09eatitself} well-typed syntax for
type theory.

TODO: list examples here and explain. Also the elimination principles
which is recursive-recursive. Talk about connection to setoidified
CwFs (I think these are explained in some Dybjer paper) and maybe the
initiality conjecture.

Reduction of IITs to inductive types is an open problem. Forsberg
showed that IITs can be reduced to inductive types in the setting of
ETT, but he only constructed the nondependent (or simple? was this the
non-recursive-recursive?) elimination principle. Hugunin \cite{jasper}
reduced an example IIT to inductive types without UIP, but also only
constructing the non-recursive-recursive eliminator. Streicher
\cite{streichersemantics}. Brunerie \cite{} constructed the initial
category with extra structure (TODO: fix this) using UIP.

Kaposi and Kov{\'a}cs defined a type theory of signatures for higher
inductive-inductive types (HIITs)
\cite{kaposi_et_al:LIPIcs:2018:9190}. This is a small type theory in
which every context encodes a valid IIT signature. For example, the
signature for natural numbers is simply the context
$(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$ and
$suc$ are simply variable names). Later they showed
\cite{Kaposi:2019:CQI:3302515.3290315} that if a subset of this theory
of signatures exists, then any finitary QIIT can be defined from
it. In this paper we further restrict the subset by not allowing
equality constructors (thus only covering IITs) but at the same time
we extend it to allow infinitary constructors. To derive the existence
of all IITs from the theory of IIT signatures we use a variant of the
CwF model \cite{Kaposi:2019:CQI:3302515.3290315} where types are
modelled not by displayed CwFs but by CwF isofibrations. We need this
extension to support infinitary constructors. Furthermore, using the
method introduced by Streicher \cite{streichersemantics} we show the
existence of the theory of IIT signatures. As opposed to Streicher, we
don't use quotients in this construction as the theory of IIT
signatures can be given only using neutral terms.

@@

Similar techique for other type theories: Streicher. Initiality
project, Brunerie. These all use quotients. Closest: Lafont TYPES
abstract 2017. Does not use quotients.

@@

The current work only concerns finitary IITs. An extension would be to also
allow infinitely branching inductive types such as W-types. This would first
require giving semantics for infinitary IITs (to our knowledge there is no
previously published semantics that we can borrow), and also giving a term model
construction analogously to finitary QIITs. These steps seem feasible. However,
it seems to be more difficult to construct the syntax of infinitary IIT
signatures without using quotients. The reason is that such syntax would not be
strictly restricted to neutral terms: we would need $\lambda$-abstraction and
$\beta\eta$-rules for infinitary $\Pi$ types, in order to allow a term model
construction for infinitary IITs. A definition of normal preterms and typing
judgments on them may still be possible, but it appears to be much more
complicated than before (the current authors have attempted this without
conclusive success).

As mentioned in Section \ref{sec:right_uniqueness}, it also remains an open
problem whether IITs are reducible to inductive types in a UIP-free setting. To
show this, we would need to construct the syntax of signatures without UIP, and
also reproduce the semantics and term model construction for IITs without UIP.

>>>>

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the James
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda.
\end{abstract}

@@

\todo{András: overall todos: introduction, section 2 examples, intro to section 3, future work finish, maybe changing big Sigma notation to (x : A) x B notation (or changing the other way?)}

Most mutual inductive types can be reduced to indexed inductive types
where the index gives the information which sort we meant. An example
is the predicates $\mathsf{isEven}$ and $\mathsf{isOdd}$ defined
mutually by the following constructors.
\begin{alignat*}{5}
  & \mathsf{isEven} && : \N\ra\Set \\
  & \mathsf{isOdd} && : \N\ra \Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,n \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,n
\end{alignat*}
These can be reduced to the following single inductive family.
\begin{alignat*}{5}
  & \mathsf{isEven?} && : \mathsf{Bool}\ra\N\ra\Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,n \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,n
\end{alignat*}

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the James Chapman style
\cite{chapman09eatitself} syntax for type theory. A minimal use case
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Con \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$
constructor of $\Con$ refers to $\Ty$ and the $\Ty$-constructor
$\mathsf{U}$ refers to $\Con$, hence the two sorts have to be defined
at the same time. Moreover, $\Ty$ is indexed over $\Con$, which
precludes the above reduction which worked for
$\mathsf{isEven}$--$\mathsf{isOdd}$ as we would get a type indexed
over itself. Another unique feature of IITs (which also shows up in
higher inductive types \cite{HoTTbook}) is that later constructors can
refer to previous constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for an IIT has two motives and one method
for each constructor. It is an example of a \emph{recursive-recursive}
function (following the nomenclature of \cite{forsberg-phd}) which
means that there are two mutually defined functions where the type of
the second function depends on the first function. The proof assistant
Agda \cite{norell07thesis} allows the definition of such functions
(even from non-IITs) and it is also currently the only proof assistant
supporting IITs. The elimination principle for $\Con$--$\Ty$ says that
given the following components,
\begin{alignat*}{5}
  & Con^M && : \Con\ra\Set \\
  & Ty^M && : Con^M\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & empty^M && : Con^M\,\mathsf{empty} \\
  & ext^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,A\ra Con^M\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,(\mathsf{U}\,\Gamma) \\
  & El^M && : (\Gamma^M:Con^M\,\Gamma)\ra\,Ty^M\,(ext^M\,\Gamma^M\,(U^M\,\Gamma^M))\,(\mathsf{El}\,\Gamma)
%  & Pi^M && : (\Gamma^M:Con^M\,\Gamma)(A^M:Ty^M\,\Gamma^M\,A)\ra Ty^M\,(ext^M\,\Gamma^M\,A^M)\ra\,Ty^M\,\Gamma^M\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
we obtain two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^M\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^M\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = empty^M \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^M\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^M\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^M\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}

As opposed to mutual inductive types, reducing IITs to inductive types
is an open problem. Forsberg \cite{forsberg-phd} showed a reduction
which only supports a simpler, non recursive-recursive elimination
principle (\todo{check}). Hugunin \cite{jasper} reduced an example IIT
to inductive types without UIP, but he also only constructed the
simpler non recursive-recursive eliminator.

The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is
a small type theory in which every context encodes a valid quotient
inductive-inductive type (QIIT) signature. For example, the signature
for natural numbers is simply the context
$(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$ and
$suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315}
showed that if this theory of signatures exists, then any finitary
QIIT can be defined from it. If we omit the type former $\mathsf{Id}$
from the theory of signatures, we get a theory of finitary IIT
signatures, and the construction of
\cite{Kaposi:2019:CQI:3302515.3290315} still works: constructing an
arbitrary IIT from the theory of IIT signatures does not need the
$\mathsf{Id}$ type former to be present.

In this paper, we construct the theory of IIT signatures only using
indexed inductive types, thus \todo{rewrite from here}

To derive the existence of all IITs from the theory of IIT signatures
we use a variant of the CwF model
\cite{Kaposi:2019:CQI:3302515.3290315} where types are modelled not by
displayed CwFs but by CwF isofibrations. We need this extension to
support infinitary constructors. Furthermore, using the method
introduced by Streicher \cite{streichersemantics} we show the
existence of the theory of IIT signatures. As opposed to Streicher, we
don't use quotients in this construction as the theory of IIT
signatures can be given only using neutral terms.

@@

(Indexed inductive types can be reduced to inductive types using a method
similar to what we use here, preterms and well-formedness predicate.)

Talk about connection to setoidified CwFs (I think these are explained
in some Dybjer paper) and maybe the initiality conjecture.

Kaposi and Kov{\'a}cs defined a type theory of signatures for higher
inductive-inductive types (HIITs)
\cite{kaposi_et_al:LIPIcs:2018:9190}.

Similar techique for other type theories: Streicher
\cite{streichersemantics}. Initiality project, Brunerie \cite{}. These
all use quotients. Closest: Lafont TYPES abstract 2017. Does not use
quotients.

@@

\subsection{Infinitary IITs}.

The current work only concerns finitary IITs. An extension would be to also
allow infinitely branching inductive types such as W-types. This would first
require giving semantics for infinitary IITs (to our knowledge there is no
previously published semantics that we can borrow), and also giving a term model
construction analogously to finitary QIITs. These steps seem feasible. However,
it seems to be more difficult to construct the syntax of infinitary IIT
signatures without using quotients. The reason is that such syntax would not be
strictly restricted to neutral terms: we would need $\lambda$-abstraction and
$\beta\eta$-rules for infinitary $\Pi$ types, in order to allow a term model
construction for infinitary IITs. A definition for normal preterms and typing
judgments on them may still be possible, but it appears to be much more
complicated than before (the current authors have attempted this without
conclusive success).

\subsection{Construction without UIP and function extensionality}
\todo{Without UIP, without funext,
for QIITs (in the setoid model).}
